The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-10 paper/ms.tex
Add dust map, use 5-10 km/s

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the \textit{Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical position of $\sim15\,\pc$ for a thin disk
orbit or $\sim120\,\pc$ for a thick disk orbit is sufficient to induce a
$25\,\%$ systematic error in the vertical action $J_z$ computed under the
assumption of an axisymmetric potential. The induced error distribution is
non-Gaussian and bimodal, with neither of the modes peaking on the null value.
Furthermore, we show that the local midplane varies by $\sim100\,\pc$ at the
Solar circle in Milky Way-mass cosmological zoom-in simulations from the FIRE
project. From observations that the mean vertical velocity of the Milky Way
disk varies by $\sim5\,\kms$ with radius, we estimate that the true midplane
variations are of order $\sim60\textup{--}90\,\pc$. Thus, current
state-of-the-art action calculations --- which assume the global and local
midplanes are the same --- are likely to include a systematic vertical offset
depending on the volume considered. Variation in the local standard of rest
induces similar issues. The variation of the midplane must be taken into
account when performing dynamical modeling across the large regions of the
disk accessible to \textit{Gaia} and future missions.

@@

The vertical frequency of the thin-disk and thick-disk orbits are $\sim
0.09\,\Myr^{-1}$ and $\sim0.06\,\Myr^{-1}$, respectively. By dimensional
analysis, and assuming a vertical velocity variation of $5\,\kms$, we
therefore expect the midplane offsets to be $\sim 57\textup{--}85\,\pc$. We
stress that this is a rough calculation, but it shows that already we see
evidence in the data from velocities for midplane offsets on the order of what
we saw in both sets of simulations.

@@

\item Assuming a vertical velocity variation of the Milky Way of
$\sim5\,\kms$, consistent with recent results from \textit{Gaia} DR2
\citep{2018A&A...616A..11G, 2019arXiv190209569F} and our results from the FIRE
simulations (Figure~\ref{fig:lsr_variations}), we estimated that the
corresponding midplane offsets are $\sim60\textup{--}90\,\pc$ by dimensional
analysis with the vertical frequencies of disk-like orbits.

>>>>

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the \textit{Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical position of $\sim15\,\pc$ for a thin disk
orbit or $\sim120\,\pc$ for a thick disk orbit is sufficient to induce a
$25\,\%$ systematic error in the vertical action $J_z$ computed under the
assumption of an axisymmetric potential. The induced error distribution is
non-Gaussian and bimodal, with neither of the modes peaking on the null value.
Furthermore, we show that the local midplane varies by $\sim100\,\pc$ at the
Solar circle in Milky Way-mass cosmological zoom-in simulations from the FIRE
project. From observations that the mean vertical velocity of the Milky Way
disk varies by $\sim5\textup{--}10\,\kms$ with radius, we estimate that the true midplane
variations are of order $\sim60\textup{--}170\,\pc$. Similar offsets are seen in three-dimensional dust maps. Thus, current
state-of-the-art action calculations --- which assume the global and local
midplanes are the same --- are likely to include a systematic vertical offset
depending on the volume considered. Variation in the local standard of rest
induces similar issues. The variation of the midplane must be taken into
account when performing dynamical modeling across the large regions of the
disk accessible to \textit{Gaia} and future missions.

@@

The vertical frequency of the thin-disk and thick-disk orbits are $\sim
0.09\,\Myr^{-1}$ and $\sim0.06\,\Myr^{-1}$, respectively. By dimensional
analysis, and assuming a vertical velocity variation of $5\textup{--}10\,\kms$, we
therefore expect the midplane offsets to be $\sim 57\textup{--}170\,\pc$. We
stress that this is a rough calculation.

Three-dimensional dust maps also offer a view into the expected variation of
the stellar disk, since dust should trace high-star formation regions.
Figure~9 of \citet{2019MNRAS.483.4277C}, Figure~1 of
\citet{2019arXiv190105971L}, and Figure~2 of \citet{2019arXiv190502734G} all
show that the midplane varies by $\sim10\degree$ at a distance of
$\sim0.75\,\kpc$, corresponding to a physical height of $\sim130\,\pc$.

Already we see evidence in the data from velocities and dust maps for midplane
offsets on the order of what we saw in both sets of simulations.

@@

\item Assuming a vertical velocity variation of the Milky Way of
$\sim5\textup{--}10\,\kms$, consistent with recent results from \textit{Gaia} DR2
\citep{2018A&A...616A..11G, 2019arXiv190209569F} and our results from the FIRE
simulations (Figure~\ref{fig:lsr_variations}), we estimated that the
corresponding midplane offsets are $\sim60\textup{--}170\,\pc$ by dimensional
analysis with the vertical frequencies of disk-like orbits. Similar offsets are seen in three-dimensional dust maps.
